Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})

### 2017-09-29 doc/turtles_statistical_learning.tex
updated analyses and manuscript to reflect different sampling approach

%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
\author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[5]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
\affil[2]{Committee on Evolutionary Biology, University of Chicago}
\affil[3]{Integrative Research Center, Field Museum of Natural History}
\affil[4]{Department of Geological Sciences, California State University -- Fullerton}
\affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}

@@

\begin{abstract}
%  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We collected landmark-based morphometric data from 366 adult \textit{E. marmorata/``E. pallida"} museum specimens. We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives. We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data. In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E. marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}. Our results indicate that there is no clear grouping of \textit{E. marmorata/``E. pallida"} based on plastron shape. In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists. Explanations for the lack of grouping in \textit{E. marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit \textit{``E. pallida"}, the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
\end{abstract}

@@

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. We use an ensemble learning approach where multiple methods are used in order to look for consensus amoung their results. We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex. In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E. marmorata} complex.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and study system}
Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters. The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}. % additional/more specific CITATIONS?

There are two major classes of machine learning method: unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous to classification and regression models \citep{Breiman1984}. Our application of the approaches used in this study illustrates only a sampling of the various methods available for fitting classification models. The classification methods used in this study were chosen because they are suited for more than two response classes.

Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}. Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes. They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results. Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.

Here, we used an ensemble of supervised machine learning methods to compare the congruence of the morphological data to different classification hypotheses. Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications. 
Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}. % MORE
%Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited. 
In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.

@@

We choose to analyze plastron shape for multiple reasons. First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat. This approach allows both museum specimens and individuals in the field to be analyzed together. Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Finally, due to this previous study a large dataset was readily avaliable.

In the case of the \textit{E. marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses. However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the molecular based hypotheses are inconsistent with whatever information is present in the morphological data. Because of unclear geographic boundaries between subgroups of \textit{E. marmorata}, we compare multiple permutations of the \citep{Spinks2010} and \citet{Spinks2014} hypotheses.

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elegans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape. We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E. marmorata} example. We predict that the \textit{E. marmorata} dataset should be of the highest complexity and our greatest challenge given \citep{Holland1992} finding that only very subtle differences existed between geographically-distinct populations.
%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first. Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge. For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \citep{Angielczyk2013a}.
% three turtle datasets
%   seven species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset is of 366 adult \textit{E. marmorata} museum specimens. These speciemens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies. Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data. For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collection archives. In cases where precise latitude and longitude information were not available we estimated for them from other locality information. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries. The taxonomic hypotheses and sub-hypotheses for the \textit{E. marorata} used here are presented in Table \ref{tab:hypotheses}.

@@

@@

\begin{table}
  \centering
  \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
  \begin{tabular}{l l l }
    \hline
    Abbreviation & Number of classes & citation \\
    \hline
    SP10.1 & 3 & \citet{Spinks2010} \\
    SP10.2 & 3 & \citet{Spinks2010} \\
    SP10.3 & 4 & \citet{Spinks2010} \\
    SP14.1 & 2 & \citet{Spinks2014} \\
    SP14.2 & 4 & \citet{Spinks2014} \\
    \hline
  \end{tabular}
  \label{tab:hypotheses}
\end{table}

Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetrical pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.

@@

\textit{Emys marmorata} is known to display sexual dimorhpism in plastron shape, particulary the presence of a plastra concavity in males \citep{Seeliger1945}. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 360).

\subsection{Supervised learning approaches}
Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between the them increases our support for that conclusion over another \citep{Hastie2009}. The supervised learning methods used here are named in Table \ref{tab:methods}. Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}. For example, multinomial logistic regression is a type of generalized linear model while random forest is an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)

\begin{table}
  \centering
  \caption{Table of the supervised learning methods used in this analysis.}
  \begin{tabular}{l l l l}
    \hline
    Method name & abbreviation & R package & citation \\
    \hline
    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
    random forests & RF & randomForest & \citet{Liaw2002} \\
    \hline
  \end{tabular}
  \label{tab:methods}
\end{table}

@@

% training and testing paradigm
%   training dataset
%     for multiple models with between 3 and 28 predictors
%     5-fold cross validation of each model
%     best model had greatest mean AUC value
%     selected model is most parsimonious model within 1 SD of ROC of best model
%   testing dataset
%     estimate class of out-of-sample testing dataset
%     compare average AUC across models for each scheme
% caret package for R
We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}. Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias. With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data are split into a training dataset and a testing dataset. The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization. In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.

@@

For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation. Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}. In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted. This is repeated for all combinations of blocks. Within each round, the predictive performance metrics are averaged across all folds. Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation. This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}. For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation. The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}. This rule is applied to ameliorate the chance of selecting an overly complex model.

@@

The results of the PCA of plastron shape in both the seven species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig. \ref{fig:other_pca}). Additionally, there appears to be both little difference in centroid size across individuals as well as little correspondence between centroid size and the PC axes.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset. Each panel corresponds to each of different proposed classification schemes analyzed as part of this study. Point color corresponds to the categories within each scheme, and the category names corresond to geographic regions. Point size is proportional to centroid size and the numbers in parantheses next to the axis labels are the percent of total variation accounted for by that axis.}
  \label{fig:emys_pca}
\end{figure}

Analysis of the differences between sexes of \textit{E. marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig. \ref{fig:sex_test}). The difference in mean shape between the sexes is very small; the sexes overlap slightly more than is expected from a null distribution based on permuting the sex-labels.

@@

\begin{table}
  \centering
  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data. Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the seven species dataset, and each of the \textit{Emys marmorata} classification schemes.}
  \begin{tabular}{l l c c}
    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
    \hline
    Replicates & & 1.11 & \\
    \hline
    Seven species & & 0.33 & 0.37 \\
    \textit{Trachemys} & & 0.76 & 0.84 \\
    \hline
    \textit{E. marmorata} & SP10.1 & 0.93 & 1.03 \\
     & SP10.2 & 0.93 & 1.03 \\
     & SP10.3 & 0.88 & 0.98 \\
     & SP14.1 & 0.98 & 1.09 \\
     & SP14.2 & 0.87 & 0.97 \\
    \hline
  \end{tabular}
  \label{tab:rep_res}
\end{table}

>>>>

%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1,2]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
\author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[5]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
\affil[2]{Committee on Evolutionary Biology, University of Chicago}
\affil[3]{Integrative Research Center, Field Museum of Natural History}
\affil[4]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
\affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}

@@

\begin{abstract}

  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We collected landmark-based morphometric data from 532 adult \textit{E. marmorata/``E. pallida''} museum specimens. We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives. We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data. In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E. marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}. Our results indicate that there is no clear ``best'' grouping of \textit{E. marmorata/``E. pallida''} based on plastron shape. In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists. Explanations for the lack of grouping in \textit{E. marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit ``E. pallida,'' the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon. 
\end{abstract}

@@

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results. We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex. In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E. marmorata} complex.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and study system}
Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality often at the expense of the interpretability of individual parameters. Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \citep{Hastie2009}. % additional/more specific CITATIONS?

There are two major classes of machine learning method: unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous to classification and regression models \citep{Breiman1984,Hastie2009}. Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models. The specific methods used in this study were chosen because they are suited for cases with more two or more response classes.

Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}. Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes. They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results. Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.

Here, we used an ensemble of supervised machine learning methods to compare the congruence of the morphological data to different classification hypotheses. Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications. 
Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}. % MORE
%Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited. 
In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified: the area under the Receiver Operating Characteristic curve \citep{Hastie2009}.

@@

We choose to analyze plastron shape for multiple reasons. First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat. This approach allows both museum specimens and individuals in the field to be analyzed together. Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Finally, due to this previous study a large dataset was readily available.

In the case of the \textit{E. marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses. However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the proposed hypotheses are inconsistent with whatever information is present in the morphological data. Because of unclear geographic boundaries between subgroups of \textit{E. marmorata}, we compare multiple permutations of the \citep{Spinks2010} and \citet{Spinks2014} hypotheses.

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elegans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape. We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E. marmorata} example. We predict that the \textit{E. marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \citep{Holland1992}.
%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first. Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge. For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \citep{Angielczyk2013a}.
% three turtle datasets
%   seven species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset is of 532 adult E. marmorata museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig. \ref{fig:map}). These specimens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. Because  \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies. Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data. For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections. In cases where precise latitude and longitude information were not available we estimated for them from other locality information. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries. The taxonomic hypotheses and sub-hypotheses for \textit{E. marmorata} used here are presented in Table \ref{tab:hypotheses} and Figure \ref{fig:map}.

For \citet{Spinks2010} we used three binning schemes. All three schemes include a class for \textit{E. marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E. pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR). The schemes differ in the assignment of samples from the San Joaquin Valley (Fig. \ref{fig:map}). Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above. In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \citet{Spinks2005}. For \citet{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis. The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \citet{Spinks2014}. Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.

@@

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_DRAFT}
  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}. Each hypothesized scheme has two or more possible classes. Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.}
  \label{fig:map}
\end{figure}

@@

\begin{table}
  \centering
    \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
    \begin{tabular}{l l l }
      \hline
      Abbreviation & Number of classes & citation \\
      \hline
      SP10.1 & 3 & \citet{Spinks2010} \\
      SP10.2 & 3 & \citet{Spinks2010} \\
      SP10.3 & 4 & \citet{Spinks2010} \\
      SP14.1 & 2 & \citet{Spinks2014} \\
      SP14.2 & 4 & \citet{Spinks2014} \\
      Morph & 2 & \citet{Spinks2010} \\
      \hline
    \end{tabular}
    \label{tab:hypotheses}
\end{table}

Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetrical pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}. All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).

@@

\textit{Emys marmorata} is known to display sexual dimorphism in plastron shape, particularly the presence of a plastra concavity in males \citep{Seeliger1945}. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532).

\subsection{Supervised learning approaches}
Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \citep{Hastie2009}. The supervised learning methods used here are named in Table \ref{tab:methods}. Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}. For example, multinomial logistic regression is a type of generalized linear model while random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)

\begin{table}
  \centering
  \caption{Table of the supervised learning methods used in this analysis.}
  \scalebox{0.9}{
  \begin{tabular}{l l l l}
    \hline
    Method name & abbreviation & R package & citation \\
    \hline
    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
    random forests & RF & randomForest & \citet{Liaw2002} \\
    \hline
  \end{tabular}}
  \label{tab:methods}
\end{table}

@@

% training and testing paradigm
%   training dataset
%     for multiple models with between 3 and 28 predictors
%     5-fold cross validation of each model
%     best model had greatest mean AUC value
%     selected model is most parsimonious model within 1 SD of ROC of best model
%   testing dataset
%     estimate class of out-of-sample testing dataset
%     compare average AUC across models for each scheme
% caret package for R
We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}. Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias. With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data are split into a training dataset and a testing dataset. The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization. For each scheme, we limited the model training and testing to only those individuals with class labels for that scheme. In this analysis, we randomly divided 80\% of samples into the training set and the remaining 20\% into the testing set.

@@

For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation. Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}. In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted. This is repeated for all combinations of blocks. Within each round, the predictive performance metrics are averaged across all folds. Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation. This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}. For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation. The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}. The purpose of this rule is to ameliorate the chances of selecting an overly complex model that will perform poorly when predicting the classes of out-of-sample data.

@@

The results of the PCA of plastron shape in both the seven species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig. \ref{fig:other_pca}).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset. Each panel corresponds to one of the six different classification schemes analyzed as part of this study (Tab. \ref{tab:hypotheses}). Point color corresponds to the categories within each scheme, and the class names correspond to geographic regions. Point size is proportional to centroid size of that specimen and the numbers in parentheses next to the axis labels are the percent of total variation accounted for along that axis.}
  \label{fig:emys_pca}
\end{figure}

Analysis of the differences between sexes of \textit{E. marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig. \ref{fig:sex_test}). The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.

@@

\begin{table}
  \centering
  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data. Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the seven species dataset, and each of the \textit{Emys marmorata} classification schemes.}
  \begin{tabular}{l l c c}
    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
    \hline
    Replicates & & 1.11 & \\
    \hline
    Seven species & & 0.33 & 0.37 \\
    \textit{Trachemys} & & 0.76 & 0.84 \\
    \hline
    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
     & SP10.2 & 1.00 & 1.11 \\
     & SP10.3 & 0.94 & 1.04 \\
     & SP14.1 & 1.01 & 1.12 \\
     & SP14.2 & 0.93 & 1.04 \\
     & Morph & 0.99 & 1.09 \\
    \hline
  \end{tabular}
  \label{tab:rep_res}
\end{table}
